### 2025-07-18 paper/quarto_aaai/paper.tex
Editorial pass up until S4.2

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be (1) plausible with
respect to the underlying data and (2) actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals ad-hoc during the
training phase to minimize the divergence between learned
representations and plausible, actionable explanations. We demonstrate
empirically and theoretically that our proposed method facilitates
training models that deliver inherently desirable explanations while
promoting robustness and preserving high predictive performance.
\end{abstract}
\section{Introduction}\label{introduction}

Today's prominence of artificial intelligence (AI) has largely been
driven by the success of representation learning with large degrees of
freedom: instead of relying on features and rules hand-crafted by
humans, modern machine learning (ML) models are tasked with learning
highly complex representations directly from the data, guided by narrow
objectives such as predictive accuracy \citep{goodfellow2016deep}. These
models tend to be so complex that humans cannot easily interpret their
decision-making.

@@

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because a narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore follow the
paradigm that models---as opposed to explanation methods---should be
held accountable for explanations that are plausible and actionable.
While previous work has shown that at least plausibility can be
indirectly achieved through existing techniques aimed at models'
generative capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

Specifically, we propose \textbf{counterfactual training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful (\emph{mature})
CEs with ground-truth data while protecting immutable features, and
capitalize on methodological links between CE and AE by penalizing the
model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
the best of our knowledge, CT represents the first venture in this
direction with promising empirical and theoretical results.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) adversarial robustness improves across the
board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values, to ensure
that individuals seeking recourse through CE face minimal costs in terms
of feature changes. Different variations of Equation~\ref{eq-general}
have been proposed in the literature to address various other desiderata
including those introduced above (faithfulness, plausibility and
actionability). Like \citet{wachter2017counterfactual}, most of these
approaches rely on gradient descent to optimize
Equation~\ref{eq-general}. For more details on various approaches tested
in this work, we refer the reader to the supplementary appendix. In the
following, we describe in detail how counterfactuals are generated and
used in CT.

@@

\begin{enumerate}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$, where $A$ is some arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ are subject to some actionability constraints.
\end{enumerate}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
CE search during training to generate on-the-fly model explanations
\(\mathbf{x}^\prime\) for training samples. We use the term
\emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
they have either exhausted all \(T\) iterations or converged in terms
reaching a pre-specified threshold, \(\tau\), for the predicted
probability of the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
components are governed through \(\lambda_{\text{div}}\),
\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
regime is sketched out in Algorithm \ref{alg-experiment}.

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence-based adaption,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

To assess outcomes with respect to actionability for non-linear models,
we look at the average costs of valid counterfactuals in terms of their
distances from factual starting points. While this an imperfect proxy of
sensitivity, we hypothesize that CT can reduce these costs by teaching
models to seek plausibility with respect to mutable features, much like
we observe in Figure~\ref{fig-poc} (panel (d) vs panel (c)). We
supplement this analysis with qualitative findings for integrated
gradients \citep{sundararajan2017ig}. Finally, for predictive
performance, we use standard metrics, such as robust accuracy estimated
on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

>>>>

\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
Deursen \and Cynthia C. S. Liem}
\date{}
\begin{document}
\maketitle
\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
user-defined mutability constraints. Much existing research has
therefore focused on developing post-hoc methods to generate
counterfactuals that meet these desiderata. In this work, we instead
hold models directly accountable for the desired end goal:
counterfactual training employs counterfactuals during the training
phase to minimize the divergence between learned representations and
plausible, actionable explanations. We demonstrate empirically and
theoretically that our proposed method facilitates training models that
deliver inherently desirable explanations while promoting robustness and
preserving high predictive performance of the models.
\end{abstract}
\section{Introduction}\label{introduction}

Today's prominence of artificial intelligence (AI) has largely been
driven by the success of representation learning with high degrees of
freedom: instead of relying on features and rules hand-crafted by
humans, modern machine learning (ML) models are tasked with learning
highly complex representations directly from the data, guided by narrow
objectives such as predictive accuracy \citep{goodfellow2016deep}. These
models tend to be so complex that humans cannot easily interpret their
decision logic.

@@

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because the narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore embrace the
paradigm that models---as opposed to explanation methods---should be
held accountable for explanations that are plausible and actionable.
While previous work has shown that at least plausibility can be
indirectly achieved through existing techniques aimed at models'
generative capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

Specifically, we propose \textbf{counterfactual training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcomes of CT compared to a conventionally trained
model. First, in panel (a), faithful and valid counterfactuals end up
near the decision boundary forming a clearly distinguishable cluster in
the target class (orange). In panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals. In panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features. By contrast,
the decision boundary in panel (d) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful CEs with
ground-truth data while protecting immutable features, and capitalize on
methodological links between CE and AE by penalizing the model's
adversarial loss on interim (\emph{nascent}) counterfactuals. To the
best of our knowledge, CT represents the first venture in this direction
with promising empirical and theoretical results.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 19\% on average; and (3) the adversarial robustness improves across
the board. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

Counterfactual explanations are typically generated by solving
variations of the following optimization problem,
\begin{equation}\phantomsection\label{eq-general}{
\begin{aligned}
\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
\end{aligned}
}\end{equation} where
\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
current model predictions for \(\mathbf{x}^{\prime}\) and the target
class (a conventional choice is cross-entropy). Finally, we use
\(\text{reg}(\cdot)\) to denote any form of regularization used to
induce certain properties on the counterfactual. In their seminal paper,
\citet{wachter2017counterfactual} propose regularizing the distance
between counterfactuals and their original factual values to ensure that
individuals seeking recourse through CE face minimal costs in terms of
feature changes. Different variations of Equation~\ref{eq-general} have
been proposed in the literature to address many desiderata including the
ones discussed above (faithfulness, plausibility and actionability).
Like \citet{wachter2017counterfactual}, most of these approaches rely on
gradient descent to optimize Equation~\ref{eq-general}. For more details
on the approaches tested in this work, we refer the reader to the
supplementary appendix. In the following, we describe in detail how
counterfactuals are generated and used in counterfactual training.

@@

\begin{itemize}
    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
\end{itemize}
and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript. 
\end{definition}

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
generated through gradient descent over \(T\) iterations as originally
proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
CE search in training to generate on-the-fly model explanations
\(\mathbf{x}^\prime\) for the training samples. We use the term
\emph{nascent} to denote interim counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
explain below, these nascent counterfactuals can be stored and
repurposed as adversarial examples. Conversely, we consider
counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
they have either exhausted all \(T\) iterations or converged by reaching
a pre-specified threshold, \(\tau\), for the predicted probability of
the target class:
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.

Formally, we propose the following counterfactual training objective to
train explainable (as in Def. \ref{def-explainability}) models:
\begin{equation}\phantomsection\label{eq-obj}{
\begin{aligned}
&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{aligned}
}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
that induces discriminative performance (e.g., cross-entropy). The
second and third terms are explained in detail below. For now, they can
be summarized as inducing explainability directly and indirectly by
penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
target class \(y^+\), and (2) the adversarial loss,
\(\text{advloss}(.)\), wrt. nascent counterfactuals
\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
of the energy terms involved in \(\text{div}(\cdot)\)
\citep{du2019implicit}. The trade-offs between these components are
adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\). The full training regime is sketched out in
Algorithm \ref{alg-experiment}.

@@

Our focus is the improvement in explainability (Def.
\ref{def-explainability}). Thus, we primarily look at the plausibility
and cost of faithfully generated counterfactuals at test time. Other
metrics, such as validity and redundancy, are reported in the
supplementary appendix. To measure the cost, we follow the standard
proxy of distances (\(\ell_1\)-norm) between factuals and
counterfactuals. For plausibility, we assess how similar CEs are to the
observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
\citet{altmeyer2024faithful},
\begin{equation}\phantomsection\label{eq-impl-dist}{
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
}\end{equation} and introduce a novel divergence-based adaptation,
\begin{equation}\phantomsection\label{eq-impl-div}{
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
the squared population maximum mean discrepancy, proposed by
\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
equal to zero if and only if the two distributions are exactly the same,
\(\mathbf{X}^\prime=\mathbf{X}^+\).

To assess outcomes with respect to actionability for non-linear models,
we look at the average costs of valid counterfactuals in terms of their
distances from factual starting points. While this an imperfect proxy of
sensitivity, we hypothesize that CT can reduce these costs by teaching
models to seek plausibility with respect to mutable features, much like
we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
supplement this analysis with qualitative findings for integrated
gradients \citep{sundararajan2017ig}. Finally, for predictive
performance, we use standard metrics, such as robust accuracy estimated
on adversarially perturbed data using FGSM
\citep{goodfellow2014explaining}.

### 2025-07-18 paper/sections/abstract.qmd
Editorial pass up until S4.2

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the user-defined mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance of the models.

### 2025-07-18 paper/sections/experiments.qmd
Editorial pass up until S4.2

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence-based adaption,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points. While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc (panel (d) vs panel (c)). We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig]. Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

>>>>

Our focus is the improvement in explainability (Def. \ref{def-explainability}). Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time. Other metrics, such as validity and redundancy, are reported in the supplementary appendix. To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals. For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the metric used by @altmeyer2024faithful,
$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}
and introduce a novel divergence-based adaptation,
$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}
where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel. The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points. While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc in panel (d) compared to (c). We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig]. Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

### 2025-07-18 paper/sections/intro.qmd
Editorial pass up until S4.2

Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with large degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. These models tend to be so complex that humans cannot easily interpret their decision-making.

@@

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.

Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful (*mature*) CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

>>>>

Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with high degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. These models tend to be so complex that humans cannot easily interpret their decision logic.

@@

Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata. This is problematic, because the narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful]. In this work, we therefore embrace the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable. While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.

Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE). @fig-poc illustrates the outcomes of CT compared to a conventionally trained model. First, in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange). In panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals. In panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features. By contrast, the decision boundary in panel (d) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature. To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals. To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.

The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the links to contrastive and robust learning. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) the adversarial robustness improves across the board. Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.
